Leaf traits drive differences in biomass partitioning among major plant functional types

### 2015-05-27 manuscript.Rnw
Moved to Ecology Letters format; fixed a reference

\usepackage[sort&compress]{natbib}
\citestyle{nature}

@@

\bibliographystyle{ieeetr}
\linenumbers

@@

\section*{First paragraph}

\textbf{The partitioning of biomass into leaves and stems is one of the most uncertain  and influential components of  global vegetation models (GVMs), affecting the residence time and storage of carbon in ecosystems, and the response of net primary production to rising CO\textsubscript{2}\citep{dekauwe2014,friend2014,ise2010}. Although GVMs typically assume that the major woody plant functional types (PFTs) differ in biomass partitioning, empirical data have been lacking. Using a new global database on biomass partitioning from over 175 studies\citep{Falster2015}, we here show a strong effect of PFT on partitioning between stems and leaves. We show that major woody PFTs do not differ in the amount of leaf area supported, but rather the amount of leaf mass per unit above ground biomass. Moreover, we find that these differences relate to systematic shifts in leaf mass per area, a leading trait within the leaf economics spectrum \citep{wrightetal2004}. In addition, we show how a second trait, leaf area per stem basal area, explains additional variation in biomass partitioning among species; however this trait does not differ systematically among major woody PFTs. Combined, these results uncover fundamental rules in how plants are constructed and allow for systematic benchmarking of biomass partitioning routines in current GVMs.
}

\section*{Main text}

@@

Moreover, we show that differences in LMF among woody PFTs arise due to differences in leaf morphology, but they converge in the amount of leaf area held on the plant (Fig. 2b-c). Previously, it was unclear whether PFTs with higher leaf biomass also had higher total leaf area. Leaf area is relevant because it drives total light interception and thus photosynthesis. Several classic studies have assumed that gymnosperms carry more leaf area than angiosperms \citep{chabothicks1982, bond1989}. We found no evidence for this, instead we found that the three PFTs maintain similar leaf area per unit above-ground biomass (the `leaf area ratio`, LAR).

@@

The finding that climate does not appreciably affect biomass partitioning is good news for models that parameterize the globe by PFT. Yet, these results contrast those of a recent study where LMF was found to increase with MAT across the globe \citep{reich2014}. One explanation for these differences is that our measurements were taken on individual plants, whereas in the other study they were on whole stands. This suggests climatic effects on stand-level biomass partitioning occur primarily by altering the size-distribution and stand density (number of individuals per unit area), rather than partitioning within individual plants. With the caveats that there is still variation between species and studies unaccounted for (Fig. S5), and that coarse climate variables such as MAT and MAP may mask small-scale climate effects on biomass partitioning, our results strongly suggest the effects of PFT are stronger than any direct climate effect on biomass partitioning within individual plants (Tables \ref{table/varpartB} and \ref{table/varpartMAPMAT}).

@@

\section{Author contributions}

R.A.D and D.S.F. conceived the ideas and methods, and co-wrote the paper. R.A.D. analysed data.

>>>>

\usepackage{natbib}
\bibliographystyle{ecol_let}

@@

\linenumbers

@@

\section*{Abstract}

% Should be 150 words - can move some to introduction
The partitioning of biomass into leaves and stems is one of the most uncertain  and influential components of  global vegetation models (GVMs), affecting the residence time and storage of carbon in ecosystems, and the response of net primary production to rising CO\textsubscript{2}\citep{dekauwe2014,friend2014,ise2010}. Although GVMs typically assume that the major woody plant functional types (PFTs) differ in biomass partitioning, empirical data have been lacking. Using a new global database on biomass partitioning from over 175 studies\citep{Falster2015}, we here show a strong effect of PFT on partitioning between stems and leaves. We show that major woody PFTs do not differ in the amount of leaf area supported, but rather the amount of leaf mass per unit above ground biomass. Moreover, we find that these differences relate to systematic shifts in leaf mass per area, a leading trait within the leaf economics spectrum \citep{wrightetal2004}. In addition, we show how a second trait, leaf area per stem basal area, explains additional variation in biomass partitioning among species; however this trait does not differ systematically among major woody PFTs. Combined, these results uncover fundamental rules in how plants are constructed and allow for systematic benchmarking of biomass partitioning routines in current GVMs.

\section*{Introduction}

@@

\section*{Results}

% Placed section cut somewhat arbitrarily!
Moreover, we show that differences in LMF among woody PFTs arise due to differences in leaf morphology, but they converge in the amount of leaf area held on the plant (Fig. 2b-c). Previously, it was unclear whether PFTs with higher leaf biomass also had higher total leaf area. Leaf area is relevant because it drives total light interception and thus photosynthesis. Several classic studies have assumed that gymnosperms carry more leaf area than angiosperms \citep{chabothicks1982, bond1989}. We found no evidence for this, instead we found that the three PFTs maintain similar leaf area per unit above-ground biomass (the `leaf area ratio`, LAR).

@@

\section*{Discussion}

% Placed section cut somewhat arbitrarily!
The finding that climate does not appreciably affect biomass partitioning is good news for models that parameterize the globe by PFT. Yet, these results contrast those of a recent study where LMF was found to increase with MAT across the globe \citep{reich2014}. One explanation for these differences is that our measurements were taken on individual plants, whereas in the other study they were on whole stands. This suggests climatic effects on stand-level biomass partitioning occur primarily by altering the size-distribution and stand density (number of individuals per unit area), rather than partitioning within individual plants. With the caveats that there is still variation between species and studies unaccounted for (Fig. S5), and that coarse climate variables such as MAT and MAP may mask small-scale climate effects on biomass partitioning, our results strongly suggest the effects of PFT are stronger than any direct climate effect on biomass partitioning within individual plants (Tables \ref{table/varpartB} and \ref{table/varpartMAPMAT}).

@@
